MG7: Configurable and scalable 16S metagenomics data analysis

### 2016-07-22 docs/mg7-paper/out/paper.tex
add preprint as a starting point for the paper, set up infrastructure

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is just an example/guide for you to refer to when submitting manuscripts to Frontiers, it is not mandatory to use Frontiers .cls files nor frontiers.tex  %
% This will only generate the Manuscript, the final article will be typeset by Frontiers after acceptance.                                                 %
%                                                                                                                                                         %
% When submitting your files, remember to upload this *tex file, the pdf generated with it, the *bib file (if bibliography is not within the *tex) and all the figures.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Version 3.1 Generated 2015/22/05 %%%
%%% You will need to have the following packages installed: datetime, fmtcount, etoolbox, fcprefix, which are normally inlcuded in WinEdt. %%%
%%% In http://www.ctan.org/ you can find the packages and how to install them, if necessary. %%%

\documentclass{frontiersSCNS} % for Science, Engineering and Humanities and Social Sciences articles
%\documentclass{frontiersHLTH} % for Health articles
%\documentclass{frontiersFPHY} % for Physics and Applied Mathematics and Statistics articles

%\setcitestyle{square}
\usepackage{url,hyperref,lineno,microtype}
\usepackage[onehalfspacing]{setspace}
\linenumbers

% fix for pandoc 1.14
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Leave a blank line between paragraphs instead of using \\

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Alexey Alekhin {et~al.}} %use et al only if is more than 1 author
\def\Authors{
 Alexey Alekhin\(^\dagger\)\,$^{1}$  Evdokim Kovach\(^\dagger\)\,$^{1}$  Marina Manrique\,$^{1}$  Pablo Pareja-Tobes\,$^{1}$  Eduardo Pareja\,$^{1}$  Raquel Tobes\,$^{1}$  and Eduardo Pareja-Tobes $^{1,*}$
}% Affiliations should be keyed to the author's name with superscript numbers and be listed as follows: Laboratory, Institute, Department, Organization, City, State abbreviation (USA, Canada, Australia), and Country (without detailed address information such as city zip codes or street names).
% If one of the authors has a change of address, list the new address below the correspondence details using a superscript symbol and use the same symbol to indicate the author in the author list.
\def\Address{$^{1}$Oh no sequences! Research Group, Era7 Bioinformatics, Granada, Spain}
% The Corresponding Author should be marked with an asterisk
% Provide the exact contact address (this time including street name and city zip code) and email of the corresponding author
\def\corrAuthor{Eduardo Pareja-Tobes}
\def\corrAddress{Oh no sequences! Research Group, Era7 Bioinformatics, Plaza Campo Verde 3, Granada, 18001, Spain}
\def\corrEmail{eparejatobes@ohnosequences.com}

\begin{document}
\onecolumn
\firstpage{1}

\title[MG7: Configurable and scalable 16S metagenomics data analysis]{MG7: Configurable and scalable 16S metagenomics data analysis -- new
methods optimized for massive cloud computing}

\author[\firstAuthorLast ]{\Authors} %This field will be automatically populated
\address{} %This field will be automatically populated
\correspondance{} %This field will be automatically populated

\extraAuth{}% If there are more than 1 corresponding author, comment this line and uncomment the next one.
%\extraAuth{corresponding Author2 \\ Laboratory X2, Institute X2, Department X2, Organization X2, Street X2, City X2 , State XX2 (only USA, Canada and Australia), Zip Code2, X2 Country X2, email2@uni2.edu}

\maketitle

\begin{abstract}

As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization.

\(^\dagger\) The first and second authors contributed equally to this
work.

\tiny
 \keyFont{ \section{Keywords:} Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
Cloud computing, NGS, Genomic big data, Microbiome, Environmental, 16S
Database } %All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\end{abstract}

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep[\citet{kodzius2015marine}]{cowan2015metagenomics}.
These new possibilities for exploring the diversity of micro-organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection
\citep[\citet{garrett2015cancer}]{zeller2014potential}. This need of
sampling across time adds more complexity to metagenomics data storage
and demands adapted algorithms to detect state variations across time as
well as idiosyncratic commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumor tissues) are required to have a sufficiently complete landscape
of the human microbiome. Moreover, re-analysis of old samples using new
tools and better reference databases might be also demanded from time to
time.

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep[\citet{harper1991record}]{harper1990extensible}.
For that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

\citep[Bio4j][]{pareja2015bio4j} is a data platform integrating data
from different resources such as UniProt or GO in a graph data paradigm.
In the assignment phase we use a subgraph containing the NCBI Taxonomy,
wrapping in Scala its Java API in a tree algebraic data type.

@@

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep[\citet{morgan2012chapter}]{segata2013computational} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database. We select the best BLAST hits (10 hits by default)
obtained for each read to do the taxonomic assignment.

@@

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity. To evaluate similarity for this first
filtering of hits we use the Expect value (by default
\(evalue \leq e^{-15}\)) that describes the number of hits one can
``expect'' to see by chance when searching a database of a particular
size. In a second filtering step we filtering those hits that are not
sufficiently good comparing them with the best one. We select the best
HSP (High Similarity Pair) per reference sequence and then choose the
best HSP (that with lowest E-value) between all the selected ones. The
bitscore of this best HSP (called S) is used as reference to filter the
rest of HSPs. All the HSPs with bitscore below the product \(p S\) are
filtered. p is a coefficient fixed by the user to define the bitscore
required, e.g.~if \(p=0.9\) and \(S=700\) the required bitscore
threshold would be \(630\). Once we have the definitive HSPs selected,
we obtain their corresponding taxonomic nodes using the taxonomic
assignments that NCBI provides for all the nt database sequences. Now we
have to analyze the topological distribution of these nodes in the
taxonomy tree: i. If all the nodes forms a line in the taxonomy tree
(are located in a not branched lineage to the tree root) we should
choose the most specific taxID as the final assignment for that read. We
call to this kind of assignment the `in line' exception (see Figure 2
right panel). ii. If not, we should search for the \emph{sensu stricto}
Lowest Common Ancestor (LCA) of all the selected taxonomic nodes (See
Figure 2 left panel). In this approach we decided to use the bitscore
for evaluating the similarity because it is a value that increases when
similarity is higher and depends a lot on the length of the HSP. Some
reads could not find sequences with enough similarity in the database
and then they would be classified as reads with no hits. Advanced
metagenomics analysis approaches \citep{huson2012microbial} have adopted
LCA assignment algorithms because it provides fine and trusted
taxonomical assignment.

@@

We decided to maintain the simpler method of Best BLAST Hit (BBH) for
taxonomic assignment because, in some cases, it can provide information
about the sequences that adds information to that obtained using LCA
algorithm. Using LCA algorithm, when some reference sequences with BLAST
alignments over the required thresholds map to a not sufficiently
specific taxID, the read can be assigned to an unspecific taxon near to
the root of the taxonomy tree. If the BBH reference sequence maps to
more specific taxa, this method, in that case, gives us useful
information.

@@

The hierarchic structure of the taxonomy of the living organisms is a
tree, and, hence, is also a graph in which each node, with the exception
of the root node, has a unique parent node. It led us to model the
taxonomy tree as a graph using the graph database paradigm. Previously
we developed Bio4j \textbf{{[}Pareja-Tobes-2015{]}}, a platform for the
integration of semantically rich biological data using typed graph
models. It integrates most publicly available data linked with sequences
into a set of interdependent graphs to be used for bioinformatics
analysis and especially for biological data. MG7 works based on the
Bio4j taxonomy module. It opens the possibility to connect the taxonomic
profiling data obtained with MG7 to all the biological knowledge
associated to each taxon. Using the information available in Bio4j for
all the proteins assigned to each taxon we are connected to all the
functional data available in Uniprot related with it.

@@

\section{Figure 1: MG7 analysis
workflow.}\label{figure-1-mg7-analysis-workflow.}

The paired reads in fastq format are merged resulting in only one
sequence per read pair. The next step is a parallelized BLASTN of every
merged sequence against the 16S reference database 16S-DB7. Then, the
mapping of the detected similar sequences in the database to the
taxonomy node to which they belong is carried out. This is done using
Bio4j that includes a module with all the NCBI taxonomy in a graph
connected with the Gene Ontology, Uniprot, and RefSeq graphs. Then the
taxonomic assignment is done for each sequence following two different
approaches: LCA and BBH, and finally the abundances corresponding to
direct and cumulative assignments for each node in percentage and
absolute counts are provided for each assignment mode.

\section{Figure 2: Lowest Common Ancestor algorithm for taxonomic
assignment.}\label{figure-2-lowest-common-ancestor-algorithm-for-taxonomic-assignment.}

The Left panel displays an example of the application of LCA algorithm
in a \emph{sensu stricto} mode. A, B, C and D represent taxonomy tree
nodes with assigned reads. Right panel displays the \emph{in line} mode
of assignment which is an exception for the \emph{sensu stricto} mode of
application of LCA algorithm. The \emph{in line} mode is used when all
the nodes are located in a line without bifurcations. In that case the
taxon assigned is the most specific (the most distant from the root).

\bibliographystyle{frontiersinSCNS_ENG_HUMS} % for Science, Engineering and Humanities and Social Sciences articles, for Humanities and Social Sciences articles please include page numbers in the in-text citations
%\bibliographystyle{frontiersinHLTH&FPHY} % for Health and Physics articles
\bibliography{refs}

>>>>

\documentclass[fleqn,10pt,lineno]{wlpeerj}

\definecolor{OldCyan}{RGB}{62,127,137}
% hyperref
\usepackage[breaklinks=true]{hyperref}
\hypersetup{
  % setpagesize=false, % page size defined by xetex
  % unicode=true, % unicode breaks when used with xetex
  pdfnewwindow,
  colorlinks,%
  citecolor=OldCyan,%
  filecolor=OldCyan,%
  linkcolor=OldCyan,%
  urlcolor=OldCyan,
  urlbordercolor=OldCyan
}
\let\OldHref\href
\renewcommand{\href}[2]{\OldHref[pdfnewwindow]{#1}{{#2}}}

% ugly fix for latest pandoc
\providecommand{\tightlist}{%
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\title{MG7: Configurable and scalable 16S metagenomics data analysis}

\author[1]{Alexey Alekhin\(^\dagger\)} \affil[1]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[2]{Evdokim Kovach\(^\dagger\)} \affil[2]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[3]{Marina Manrique} \affil[3]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[4]{Pablo Pareja-Tobes} \affil[4]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[5]{Eduardo Pareja} \affil[5]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[6]{Raquel Tobes} \affil[6]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}
\author[7]{Eduardo Pareja-Tobes} \affil[7]{\emph{\href{http://ohnosequences.com}{oh no sequences!}} research group,
\href{http://www.era7bioinformatics.com}{Era7 bioinformatics}}

\keywords{ Metagenomics, 16S, Bacterial diversity profile, Bio4j, Graph databases,
 Cloud computing, NGS, Genomic big data, Microbiome, Environmental, 16S
 Database}

% PeerJ doesn't like numbered sections
\setcounter{secnumdepth}{0}

% \renewcommand\refname{}

\begin{abstract}
As part of the Cambrian explosion of omics data, metagenomics brings to
the table a specific, defining trait: its social essence. The
\emph{meta} prefix exerts its influence, with multitudes manifesting
themselves everywhere; from samples to data analysis, from actors
involved to (present and future) applications. Of these dimensions, data
analysis is where needs lay further from what current tools provide. Key
features are, among others, scalability, reproducibility, data
provenance and distribution, process identity and versioning. These are
the goals guiding our work in MG7, a 16S metagenomics data analysis
system. The basic principle is a new approach to data analysis, where
configuration, processes, or data locations are static, type-checked and
subject to the standard evolution of a well-maintained software project.
Cloud computing, in its Amazon Web Services incarnation, when coupled
with these ideas, produces a robust, safely configurable, scalable tool.
Processes, data, machine behaviors and their dependencies are expressed
using a set of libraries which bring as much as possible checking and
validation to the type level, without sacrificing expressiveness.
Together they form a toolkit for defining scalable cloud-based workflows
composed of stateless computations, with a static reproducible
specification of dependencies, behavior and wiring of all steps. The
modeling of taxonomy data is done using Bio4j, where the new paradigm of
graph databases allows for both a simple expression of taxonomic
assignment tasks and the calculation of taxa abundance values
considering the hierarchic structure of the taxonomy tree. MG7 includes
a new 16S reference database, \emph{16S-DB7}, built with a flexible and
sustainable update system, and the possibility of project-driven
personalization. \(^\dagger\) The first and second authors contributed
equally to this work.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

@@

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies \citep{oulas2015metagenomics}, especially in human
health (diagnostics, treatments, drug response or prevention)
\citep{bikel2015combining}. We should also mention what could be called
the microbiome explosion: all kind of microbiomes (gut, mouth, skin,
urinary tract, airway, milk, bladder) are now routinely sequenced in
different conditions of health and disease, or after different
treatments. The impact of Metagenomics is also being felt in
environmental sciences \citep{ufarte2015discovery}, crop sciences, the
agrifood sector \citep{coughlan2015biotechnological} and biotechnology
in general \citep{cowan2015metagenomics} \citep{kodzius2015marine}.
These new possibilities for exploring the diversity of micro-organisms
in the most varied environments are opening new research areas, and
drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling from the same patient
\citep{faust2015metagenomics} along several weeks (or years) is being
used for the follow up of some intestinal pathologies, for studying the
evolution of the gut microbiome after antibiotic treatment, or for colon
cancer early detection \citep{zeller2014potential}
\citep{garrett2015cancer}. This need of sampling across time adds more
complexity to metagenomics data storage and demands adapted algorithms
to detect state variations across time as well as idiosyncratic
commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumor tissues) are required to have a sufficiently complete landscape
of the human microbiome. Moreover, re-analysis of old samples using new
tools and better reference databases might be also demanded from time to
time.

@@

\href{https://github.com/ohnosequences/datasets}{Datasets} is a Scala
library developed by the first and last authors with the goal of being a
Scala-embedded mini-language for datasets and their locations.
\textbf{Data} is represented as type-indexed fields: keys are modeled as
singleton types, and values correspond to what could be called a
denotation of the key: a value of type \texttt{Location} tagged with the
key type. Then a \textbf{Dataset} is essentially a collection of data,
which are guaranteed statically to be different through type-level
predicates, making use of the value--type correspondence which can be
established through singleton types and implicits. A dataset location is
then just a list of locations formed by locations of each dataset key.
All this is based on what could be described as an embedding in Scala of
an extensible record system with concatenation on disjoint labels, in
the spirit of \citep{harper1990extensible} \citep{harper1991record}. For
that \emph{Datasets} uses
\href{https://github.com/ohnosequences/cosas/}{ohnosequences/cosas}
library.

@@

Bio4j \citep{pareja2015bio4j} is a data platform integrating data from
different resources such as UniProt or GO in a graph data paradigm. In
the assignment phase we use a subgraph containing the NCBI Taxonomy,
wrapping in Scala its Java API in a tree algebraic data type.

@@

\begin{figure}[htbp]
\centering
\includegraphics{./Figure-1.jpg}
\caption{\textbf{MG7 analysis workflow.} The paired reads in fastq
format are merged resulting in only one sequence per read pair. The next
step is a parallelized BLASTN of every merged sequence against the 16S
reference database 16S-DB7. Then, the mapping of the detected similar
sequences in the database to the taxonomy node to which they belong is
carried out. This is done using Bio4j that includes a module with all
the NCBI taxonomy in a graph connected with the Gene Ontology, Uniprot,
and RefSeq graphs. Then the taxonomic assignment is done for each
sequence following two different approaches: LCA and BBH, and finally
the abundances corresponding to direct and cumulative assignments for
each node in percentage and absolute counts are provided for each
assignment mode.}
\end{figure}

@@

The second step is to search for similar 16S sequences in our 16S-DB7
database. The taxonomic assignment for each read is based on BLASTN of
each read against the 16S database.Assignment based on direct similarity
of each read one by one compared against a sufficiently wide database is
considered in different reviews of metagenomics analysis methodologies
\citep{segata2013computational} \citep{morgan2012chapter} as a very
exhaustive method for assignment. Some methods of assignment compare the
sequences only against the 16S genes from available complete bacterial
genomes or avoid computational cost clustering or binning the sequences
first, and then doing the assignments only for the representative
sequence of each cluster. MG7 carries out an exhaustive comparison of
all the reads under analysis and it does not applies any binning
strategy. Every read is specifically compared with all the sequences of
the 16S database. We select the best BLAST hits (10 hits by default)
obtained for each read to do the taxonomic assignment.

@@

\begin{figure}[htbp]
\centering
\includegraphics{./Figure-2.jpg}
\caption{\textbf{Lowest Common Ancestor algorithm for taxonomic
assignment.} The Left panel displays an example of the application of
LCA algorithm in a \emph{sensu stricto} mode. A, B, C and D represent
taxonomy tree nodes with assigned reads. Right panel displays the
\emph{in line} mode of assignment which is an exception for the
\emph{sensu stricto} mode of application of LCA algorithm. The \emph{in
line} mode is used when all the nodes are located in a line without
bifurcations. In that case the taxon assigned is the most specific (the
most distant from the root).}
\end{figure}

@@

For each read, first, we select the BEST BLAST HITs (by default 10 Hits)
over a threshold of similarity. To evaluate similarity for this first
filtering of hits we use the Expect value (by default
\(evalue \leq e^{-15}\)) that describes the number of hits one can
``expect'' to see by chance when searching a database of a particular
size. In a second filtering step we filtering those hits that are not
sufficiently good comparing them with the best one. We select the best
HSP (High Similarity Pair) per reference sequence and then choose the
best HSP (that with lowest E-value) between all the selected ones. The
bitscore of this best HSP (called S) is used as reference to filter the
rest of HSPs. All the HSPs with bitscore below the product \(p S\) are
filtered. p is a coefficient fixed by the user to define the bitscore
required, e.g.~if \(p=0.9\) and \(S=700\) the required bitscore
threshold would be \(630\). Once we have the definitive HSPs selected,
we obtain their corresponding taxonomic nodes using the taxonomic
assignments that NCBI provides for all the nt database sequences. Now we
have to analyze the topological distribution of these nodes in the
taxonomy tree: i. If all the nodes forms a line in the taxonomy tree
(are located in a not branched lineage to the tree root) we should
choose the most specific taxID as the final assignment for that read. We
call to this kind of assignment the `in line' exception (see Figure 2
right panel). ii. If not, we should search for the \emph{sensu stricto}
Lowest Common Ancestor (LCA) of all the selected taxonomic nodes (See
Figure 2 left panel). In this approach we decided to use the bitscore
for evaluating the similarity because it is a value that increases when
similarity is higher and depends a lot on the length of the HSP. Some
reads could not find sequences with enough similarity in the database
and then they would be classified as reads with no hits. Advanced
metagenomics analysis approaches \citep{huson2012microbial} have adopted
LCA-based assignment algorithms because it provides fine and trusted
taxonomical assignment.

@@

We decided to maintain the simpler method of Best BLAST Hit (BBH) for
taxonomic assignment because, in some cases, it can provide information
about the sequences that adds information to that obtained using the LCA
algorithm. With the LCA algorithm, when some reference sequences with
BLAST alignments over the required thresholds map to a not sufficiently
specific taxID, the read can be assigned to an unspecific taxon near to
the root of the taxonomy tree. If the BBH reference sequence maps to
more specific taxa, this method, in that case, gives us useful
information.

@@

The hierarchic structure of the taxonomy of the living organisms is a
tree, and, hence, is also a graph in which each node, with the exception
of the root node, has a unique parent node. It led us to model the
taxonomy tree as a graph using the graph database paradigm. Previously
we developed Bio4j \citep{pareja2015bio4j}, a platform for the
integration of semantically rich biological data using typed graph
models. It integrates most publicly available data linked with sequences
into a set of interdependent graphs to be used for bioinformatics
analysis and especially for biological data. MG7 works based on the
Bio4j taxonomy module. It opens the possibility to connect the taxonomic
profiling data obtained with MG7 to all the biological knowledge
associated to each taxon. Using the information available in Bio4j for
all the proteins assigned to each taxon we are connected to all the
functional data available in Uniprot related with it.

@@

\bibliography{./out/../refs}
